Revisiting the hydrodynamic modulation of short surface waves by longer waves

### 2025-03-10 paper.tex
Addressing reviews

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the wave crest and action conservation laws and derive
higher-order steady solutions for the change of short-wave wavenumber, action,
and gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
The short-wave steepness modulation is attributed to be 5/8 due to the
wavenumber, 1/4 due to the wave action, and 1/8 due to the effective gravity.
Examining the homogeneity and stationarity requirements of the wave action
conservation equation reveals that stationarity is more easily violated than
homogeneity, and that the short-wave action is strongly conserved ($>$99\%) only
for very small amplitude long waves, and weakly conserved ($>$90\%) for steep
long waves.
We further examine the results of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs for homogeneous initial
conditions as a special case and does not generally occur, for example in more
realistic long-wave groups.
The proposed steady solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx$0.2.
Except for a subset of initial conditions, the solutions to the fully nonlinear
crest-action conservation equations are mostly steady in the reference frame of
the long waves.
\end{abstract}

@@

\begin{equation}
\label{eq:gravity_modulation_linear_no_curvature_mean_level}
\frac{\widetilde{g_e}}{g} = 1 - \varepsilon_L \cos{\psi}
\end{equation}
which is the form used by \citet{longuet1960changes} and
\citet{peureux2021unsteady}, for example.

@@

\begin{equation}
\label{eq:wave_amplitude_modulation_order1}
\dfrac{\widetilde{a}}{a} = 
  \left( \dfrac{g}{\widetilde{g}} \right)^{\frac{1}{4}}
  \left( 1 + \varepsilon_L \right)^{\frac{3}{4}}
  \approx \frac{(1 + \varepsilon_L)^{\frac{3}{4}}}{(1 - \varepsilon_L)^{\frac{1}{4}}}
  \approx 1 + \varepsilon_L
\end{equation}

@@

Finally, the short-wave steepness modulation follows from (\ref{eq:k_short_exact})
and (\ref{eq:wave_amplitude_modulation}):

\begin{equation}
\label{eq:ak_modulation}
\dfrac{\widetilde{ak}}{ak} = 
  \left( \dfrac{\widetilde{k}}{k} \right)^{\frac{5}{4}}
  \left( \dfrac{\widetilde{N}}{N} \right)^{\frac{1}{2}}
  \left( \dfrac{g}{\widetilde{g}} \right)^{\frac{1}{4}}
\end{equation}
The short-wave steepness modulation at the long-wave crests is thus largely
controlled by the wavenumber modulation (5/8, or, 62.5\%), followed by the
convergence of wave action (1/4, or, 25\%), and by least amount by the
effective gravity reduction (1/8, or, 12.5\%) (Fig. \ref{fig:modulation_contributors}).
(\ref{eq:ak_modulation}) and Fig. \ref{fig:modulation_contributors} demonstrate
that more than half of the steepness modulation can be captured by the
wavenumber modulation alone.
Further, considering the wavenumber and wave action modulations, while
neglecting the gravity modulation, captures $\approx$88\% of the steepness
modulation at the long-wave crests.
This result may be useful for the interpretation of remote sensing products
that resolve some but not all of the short-wave modulation effects.
Finally, the relative contributions of different modulation factors remain
mostly constant as $\varepsilon_L$ increases, with only a mild increase of the
effective gravity modulation contribution, and mild increases of the
wavenumber and wave action modulation contributions.

@@

The maximum modulations of the short-wave wavenumber, amplitude, and steepness
from the analytical and numerical solutions, are shown on Figs.
\ref{fig:numerical_vs_analytical_k},
\ref{fig:numerical_vs_analytical_a}, and
\ref{fig:numerical_vs_analytical_ak}, respectively, as function of the long-wave
steepness $\varepsilon_L$.
Overall, the numerical solutions, based on the linear and the third-order Stokes
long waves alike, begin to diverge from the analytical solution at
$\varepsilon_L \approx 0.15$.
The numerical solution using the third-order Stokes long wave is quantitatively
very similar to the fully-nonlinear solutions of \citet{longuet1987propagation}
in wavenumber modulation, however with somewhat lower amplitude and steepness
modulations.
Using the velocity and gravity properties of a fully-nonlinear wave, based on
the approach by \citet{clamond2018accurate}, the numerical solutions diverge
from those of the third-order Stokes wave only at very large steepnesses
($\varepsilon_L > 0.3$).
Using a fully-nonlinear form of the long-wave thus does not collapse the
difference between the solutions here and those of \citet{longuet1987propagation}.
For example, at $\varepsilon_L = 0.4$, the steepness modulation in our solution
is 6.6, whereas \citet{longuet1987propagation} finds it to be 7.8.
A possible reason for this difference is that the solutions by
\citet{longuet1987propagation} are stationary (see his \S2), whereas in our
numerical model we solve the prognostic nonlinear equations for the short-wave
wavenumber and action.
As we will see in the next section, and as was suggested by \citet{peureux2021unsteady},
they are often, but not always, mostly steady, and can be strongly unsteady in
special cases.
Nevertheless, the overall steepness modulation between the two numerical
approaches are remarkably similar, and either solution would cause the
short-waves of even low steepness to break \citep{banner1993wave}.
Where the two modulation solutions differ (at $\varepsilon_L > 0.2$),
either solution would lead to short-wave breaking after one long-wave period.
This suggests that in the cases where the short-waves would survive the
modulation and persist over multiple long-wave periods, the two solutions are
effectively equivalent.

@@

\subsection{Short-wave modulation by long-wave groups}
\label{subsection:wave_groups}

We now proceed to examine the numerical solutions with the long-wave forcing
applied as a long-wave group.
The wave group is simulated as a non-dispersive train of long waves whose
amplitude scales with $\sin[\pi t / (n T_L)]$, where $n$ is the number of
long waves in the group.
The elevation of the long waves thus gradually ramps up from rest to $a_L$ at
$t = n T_L / 2$, after which it gradually decreases toward zero.
In this simulation, $n = 10$ and $\varepsilon_L = 0.1$ as in the previous two
simulations.
The modulations of short-wave wavenumber, amplitude, and steepness in response
to such long-wave group is shown in Fig. \ref{fig:modulation_3panel_groups}.
Like in the case of a linear ramp, the modulation pattern here is locked in to
the long-wave crest and peaks at approximately 1.2 for short-wave steepness of 0.1.
After the peak of the long-wave group passes, the short-wave modulation recedes
back toward the resting state.

@@

In this paper we revisited the problem of short-wave modulation by long waves
from a wave action conservation perspective.
Using the linearised wave crest and action conservation equations coupled with
nonlinear solutions for the effective gravity, we derived steady analytical
solutions for the modulation of short wave wavenumber, amplitude, steepness,
intrinsic frequency, and phase speed.
The latter two have not been previously examined in the context of short-wave
modulation, but may be important for correctly calculating the wind input into
short waves, which is proportional to the wind speed relative to the wave
celerity.
The approximate steady solutions yield higher wavenumber and amplitude
modulation than that of \citet{longuet1960changes}, and somewhat lower than 
the numerical solutions of \citet{longuet1987propagation} and
\citet{zhang1990evolution}.
This is mainly due to the analytical solution requiring the crest and action
conservation equations to be linearised, and the nonlinearity arising due
to the evaluation of short waves on the surface of the long wave (as opposed to
the mean water level) and due to nonlinear effective gravity.
In the steady solutions, the short-wave steepness modulation is dominated by
the wavenumber modulation (5/8), followed by the wave action modulation (1/4),
and is least affected by the effective gravity modulation (1/8).

>>>>

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the wave crest and action conservation laws and derive
higher-order steady solutions for the change of short-wave wavenumber, action,
and gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
The short-wave steepness modulation is attributed to be 5/8 due to the
wavenumber, 1/4 due to the wave action, and 1/8 due to the effective gravity.
Examining the homogeneity and stationarity requirements of the wave action
conservation equation reveals that stationarity is more easily violated than
homogeneity, and that the short-wave action is strongly conserved ($>$99\%) only
for very small-amplitude long waves, and weakly conserved ($>$90\%) for steep
long waves.
We further examine the results of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs for homogeneous initial
conditions as a special case and does not generally occur, for example in more
realistic long-wave groups.
The proposed steady solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx$0.2.
Except for a subset of initial conditions, the solutions to the fully nonlinear
crest-action conservation equations are mostly steady in the reference frame of
the long waves.
\end{abstract}

@@

\begin{equation}
\label{eq:gravity_modulation_linear_no_curvature_mean_level}
\frac{\widetilde{g_e}}{g} = 1 - \varepsilon_L \cos{\psi}
\end{equation}
which is the form used by \citet{longuet1960changes} and
\citet{peureux2021unsteady}, for example.
The Lagrangian correction in (\ref{eq:gravity_modulation_lagrangian}) is:

\begin{equation}
\label{eq:gravity_modulation_lagrangian_correction}
U \frac{\partial W}{\partial x} = g \varepsilon_L^2 \cos^2{\psi} e^{2 \varepsilon_L \cos{\psi}}
+ \mathcal{O}(\varepsilon_L^3)
\end{equation}
and so it becomes important only for large $\varepsilon_L$.

@@

\begin{equation}
\label{eq:wave_amplitude_modulation_order1}
\dfrac{\widetilde{a}}{a} = 
  \left( \dfrac{\widetilde{g}}{g} \right)^{-\frac{1}{4}}
  \left( 1 + \varepsilon_L \right)^{\frac{3}{4}}
  \approx \frac{(1 + \varepsilon_L)^{\frac{3}{4}}}{(1 - \varepsilon_L)^{\frac{1}{4}}}
  \approx 1 + \varepsilon_L
\end{equation}

@@

Finally, the short-wave steepness modulation follows from (\ref{eq:wave_amplitude_modulation}):

\begin{equation}
\label{eq:ak_modulation}
\dfrac{\widetilde{ak}}{ak} = 
  \left( \dfrac{\widetilde{k}}{k} \right)^{\frac{5}{4}}
  \left( \dfrac{\widetilde{N}}{N} \right)^{\frac{1}{2}}
  \left( \dfrac{\widetilde{g}}{g} \right)^{-\frac{1}{4}}
\end{equation}
The short-wave steepness modulation at the long-wave crests is thus largely
controlled by the wavenumber modulation (5/8, or, 62.5\%), followed by the
convergence of wave action (1/4, or, 25\%), and by least amount by the
effective gravity reduction (1/8, or, 12.5\%) (Fig. \ref{fig:modulation_contributors}).
(\ref{eq:ak_modulation}) and Fig. \ref{fig:modulation_contributors} demonstrate
that more than half of the steepness modulation can be captured by the
wavenumber modulation alone.
Further, considering the wavenumber and wave action modulations, while
neglecting the gravity modulation, captures $\approx$88\% of the steepness
modulation at the long-wave crests.
This result may be useful for the interpretation of remote sensing products
that resolve some but not all of the short-wave modulation effects.
Finally, the relative contributions of different modulation factors remain
mostly constant as $\varepsilon_L$ increases, with only a mild decrease of the
effective gravity modulation contribution, and mild increases of the
wavenumber and wave action modulation contributions.

@@

The maximum modulations of the short-wave wavenumber, amplitude, and steepness
from the analytical and numerical solutions, are shown on Figs.
\ref{fig:numerical_vs_analytical_k},
\ref{fig:numerical_vs_analytical_a}, and
\ref{fig:numerical_vs_analytical_ak}, respectively, as function of the long-wave
steepness $\varepsilon_L$.
Overall, the numerical solutions, based on the linear and the third-order Stokes
long waves alike, begin to diverge from the analytical solution at
$\varepsilon_L \approx 0.15$.
The numerical solution using the third-order Stokes long wave is quantitatively
very similar to the fully-nonlinear solutions of \citet{longuet1987propagation}
in wavenumber modulation, however with somewhat lower amplitude and steepness
modulations.
Using the velocity and gravity properties of a fully-nonlinear wave, based on
the approach by \citet{clamond2018accurate}, the numerical solutions diverge
from those of the third-order Stokes wave only at very large steepnesses
($\varepsilon_L > 0.3$).
Using a fully-nonlinear form of the long-wave thus does not collapse the
difference between the solutions here and those of \citet{longuet1987propagation}.
For example, at $\varepsilon_L = 0.4$, the steepness modulation in our solution
is 6.6, whereas \citet{longuet1987propagation} finds it to be 7.8.
A possible reason for this difference is that the solutions by
\citet{longuet1987propagation} are stationary (see his \S2), whereas in our
numerical model we solve the prognostic nonlinear equations for the short-wave
wavenumber and action.
As we will see in the next section, and as was suggested by \citet{peureux2021unsteady},
they are often, but not always, mostly steady, and can be strongly unsteady in
special cases.
Nevertheless, the overall steepness modulation between the two numerical
approaches are remarkably similar, and either solution would cause the
short-waves of moderate steepness to break \citep{banner1993wave}.
Specifically, where the two modulation solutions differ ($\varepsilon_L \gtrsim 0.3$),
the steepness modulation approaches and exceeds 3.
Tripling the short-wave steepness of even $\varepsilon = 0.1$ would bring the
wave train near the observed breaking thresholds of $\varepsilon \gtrsim 0.32$
\citep{perlin2013breaking}, although, recent measurements show evidence of
significantly higher breaking thresholds
\citep{toffoli2010maximum,mcallister2024three}, and even exceeding the
geometric Stokes limit of 0.44.
Nevertheless, in the cases where the short-waves would survive the
modulation and persist over multiple long-wave periods, the two solutions are
effectively equivalent.

@@

\subsection{Modulation by long waves of varying amplitude}
\label{subsection:wave_groups}

We now proceed to examine the numerical solutions of short-wave modulation by
long waves whose amplitude varies with time.
The amplitude variation aims to mimic the behaviour of long-wave groups, a
phenomenon that is ubiquitous in the ocean due to the dispersive nature of
surface gravity waves, as well as due to the fact that wind-generated waves
evolve to have a broad-banded spectra.
For simplicity of implementation and interpretation, 
the wave group here is simulated as a non-dispersive train of long waves whose
amplitude scales with $\sin[\pi t / (n T_L)]$, where $n$ is the number of
long waves in the group.
The elevation of the long waves thus gradually ramps up from rest to $a_L$ at
$t = n T_L / 2$, after which it gradually decreases toward zero.
In this simulation, $n = 10$ and $\varepsilon_L = 0.1$ as in the previous two
simulations.
The modulations of short-wave wavenumber, amplitude, and steepness in response
to such long-wave group is shown in Fig. \ref{fig:modulation_3panel_groups}.
Like in the case of a linear ramp, the modulation pattern here is locked in to
the long-wave crest and peaks at approximately 1.2 for short-wave steepness of 0.1.
After the peak of the long-wave group passes, the short-wave modulation recedes
back toward the resting state.

@@

In this paper we revisited the problem of short-wave modulation by long waves
from a wave action conservation perspective.
Using the linearised wave crest and action conservation equations coupled with
nonlinear solutions for the effective gravity, we derived steady analytical
solutions for the modulation of short wave wavenumber, amplitude, steepness,
intrinsic frequency, and phase speed.
The latter two were previously examined in the context of short-wave
modulation \citep{longuet1962resonant,longuet1962phase}, and may be important
for correctly calculating the wind input into short waves, which is proportional
to the wind speed relative to the wave celerity.
The approximate steady solutions yield higher wavenumber and amplitude
modulation than that of \citet{longuet1960changes}, and somewhat lower than 
the numerical solutions of \citet{longuet1987propagation} and
\citet{zhang1990evolution}.
This is mainly due to the analytical solution requiring the crest and action
conservation equations to be linearised, and the nonlinearity arising due
to the evaluation of short waves on the surface of the long wave (as opposed to
the mean water level) and due to nonlinear effective gravity.
In the steady solutions, the short-wave steepness modulation is dominated by
the wavenumber modulation (5/8), followed by the wave action modulation (1/4),
and is least affected by the effective gravity modulation (1/8).
